---
title: "Project 2"
author: "Shayne Cassidy, Sarah Nelson, Sarah Winston Nathan"
date: "12-6-2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
require("knitr")
#datadir <- "/Users/shaynecassidy/Desktop/4021/Data/AirQualityUCI"
#sourcedir <-"/Users/shaynecassidy/Desktop/4021/RCode"

# sourcedir <- "C:/Users/swnathan/Documents/4y1s/sys4021/InClassR/RCode"
# datadir <- "C:/Users/swnathan/Documents/4y1s/sys4021/InClassR/Data/AirQualityUCI"
sourcedir <- "C:/Users/student/SYS4021/RCode"
datadir <- "C:/Users/student/SYS4021/Data/AirQualityUCI/"
opts_knit$set(root.dir = sourcedir)
library(forecast)
library(mtsdi)
library(MTS)
```

# Load data and impute missing values
```{r, warning=FALSE}
setwd(datadir)
airquality = read.csv('AirQualityUCI.csv')

# replace -200 with NA
airquality[airquality == -200] <- NA

# convert integer type to numeric
intcols = c(4,5,7,8,9,10,11,12)
for(i in 1:length(intcols)){
  airquality[,intcols[i]] <- as.numeric(airquality[,intcols[i]])
}

setwd(sourcedir)

# create new data frame with just CO and NO2
AQdata = airquality[,c(3,10)]

# impute missing air quality data
f <- ~ CO.GT. + NO2.GT.
t <- c(seq(1,dim(AQdata)[1],1))
i <- mnimput(f, AQdata, eps=1e-3, ts=TRUE, method='gam', 
             ga.control=list(formula=paste(names(AQdata)[c(1:3)],'~ns(t,2)')))

# set airquality to imputed data
AQdata <- i$filled.dataset

# aggregate to daily maxima for model building
dailyAQ <- aggregate(AQdata, by=list(as.Date(airquality[,1],"%m/%d/%Y")), FUN=max)

# remove last 7 days
dailyAQ <- dailyAQ[1:(dim(dailyAQ)[1]-7),]
```

### Part 1: Building Univariate Time Series Models
```{r}
AQ.CO <- dailyAQ$CO.GT.
#AQ.CO <- AQdata$CO.GT.
AQ.NO2 <- dailyAQ$NO2.GT.
#AQ.NO2<-AQdata$NO2.GT.

CO.ts<-ts(AQ.CO)
NO2.ts<-ts(AQ.NO2)

plot(CO.ts)
plot(NO2.ts)
```

# Part A: Seasonality 
```{r}
acf(CO.ts)
acf(NO2.ts)

pg.CO <- spec.pgram(CO.ts,spans=9,demean=T,log='no')
pg.NO2 <- spec.pgram(NO2.ts,spans=9,demean=T,log='no')
```
The spikes in both periodagrams at repeated frequencies indicates seasonality present.

```{r}
# What are the periods of the next biggest peaks?
# sort spectrum from largest to smallest and find index
sorted.spec <- sort(pg.CO$spec, decreasing=T, index.return=T)
names(sorted.spec)

# corresponding periods
sorted.omegas <- pg.NO2$freq[sorted.spec$ix]
sorted.Ts <- 1/pg.NO2$freq[sorted.spec$ix]

# look at first 20
sorted.omegas[1:20]
sorted.Ts[1:20]
# evens around 7
period<-7 
```

# Part B: Trends

Build a new model, CO.trend which predicts CO.ts based on the time variable
```{r}
time<-c(1:(length(CO.ts)))
CO.trend<-lm(CO.ts ~ time)
NO2.trend<-lm(NO2.ts ~ time)

summary(CO.trend)
summary(NO2.trend)
```
Here we built two new models, CO.trend and No2.trend, that both model the trend components. 
For CO.trend, the p-value is 0.00097, and for NO2.trend, the p-value is <2.2e-16. Therefore, the trend 
component is significant in both models and must be considered.

# Plot CO.trend model
```{r}
{plot(time, CO.ts, type = "l")
abline(CO.trend, col = "red")}
```
As seen in the plot of the CO.trend model, we can see that there is a clear upward trend line, which 
supports the results of our statistical test. 
The adjusted R^2 for the model CO.trend is 0.02558. 

# Plot NO2.trend model
```{r}
{plot(time, NO2.ts, type = "l")
abline(NO2.trend, col = "red")}
```
As seen in the plot of the NO2.trend model, we can see that there is a clear upward trend line.
From the naked eye, the slope seems more drastic than with the trend line from CO.trend model. 
This supports the results of our statistical test. 
The adjusted R^2 for the model NO2.trend is 0.3031.

# Model diagnostics for CO.trend --> talk about why we don't do a log transform here!!
```{r}
par(mfrow=c(2,2))
plot(CO.trend, labels.id = NULL)
par(mfrow=c(1,1))
```
Residuals versus fitted plot: Does not violate assumptions. The mean is about zero and there seems to be constant variance. There are a few outliers.
Q-Q plot: The fit to the line is fairly solid, thus no drastic violation of assumptions. The Q-Q plot could be improved.
Scale-location: Does not violate assumptions. The mean is about zero and there seems to be constant variance. There are a few outliers.
Residuals versus leverage: No clear influential points with regards to Cook's distance.

Based on these diagnostics plots, it does not appear that we need to perform any type of transformation (i.e. Log transform).

# Model diagnostics for NO2.trend
```{r}
par(mfrow=c(2,2))
plot(NO2.trend, labels.id = NULL)
par(mfrow=c(1,1))
```
Residuals versus fitted plot: Does not violate assumptions. The mean is about zero and there seems to be constant variance.
Q-Q plot: The fit to the line is fairly solid, thus no drastic violation of assumptions. As seen by the naked eye, the Q-Q plot for N02.trend is a little better than that of CO.trend.
Scale-location: Does not violate assumptions. The mean is about zero and there seems to be constant variance. There are a few outliers.
Residuals versus leverage: No clear influential points with regards to Cook's distance.

Based on these diagnostics plots, it does not appear that we need to perform any type of transformation (i.e. Log transform).

# Add seasonality component to CO.trend
Because the seasonality component was significant, we added a seasonality component to CO.trend. We decided to use a period of 7 because the peak frequency of the periodogram is about 0.14, and 1/0.14 = 7. This implies a weekly period. 
```{r}
CO.trend.seasonal <- lm(CO.ts[time] ~ time + sin(2*pi*time/7) + cos(2*pi*time/7))
summary(CO.trend.seasonal)
```
The p-value for this model is 5.027e-13.
The adjusted R^2 for the model CO.trend.seasonal is 0.1399. 

# add seasonality component to NO2.trend
Because the seasonality component was significant, we added a seasonality component to N02.trend. We decided to use a period of 7 because the peak frequency of the periodogram is about 0.14, and 1/0.14 = 7. This implies a weekly period. 
```{r}
NO2.trend.seasonal <- lm(NO2.ts[time] ~ time + sin(2*pi*time/7) + cos(2*pi*time/7))
summary(NO2.trend.seasonal)
```
The p-value for this model is 2.2e-16. 
The adjusted R^2 for the model N02.trend.seasonal is 0.3525 

# Model diagnostics for CO.trend.seasonal
```{r}
par(mfrow=c(2,2))
plot(CO.trend.seasonal, labels.id = NULL)
par(mfrow=c(1,1))
```
Residuals versus fitted plot: Does not violate assumptions. The mean is about zero and there seems to be constant variance. 
Q-Q plot: The fit to the line is fairly solid, thus no drastic violation of assumptions. 
Scale-location: Does not violate assumptions. The mean is about zero and there seems to be constant variance. 
Residuals versus leverage: No clear influential points with regards to Cook's distance. 
The spread of points above and below the mean line in the residuals versus fitted plot and scale-location plots have improved from the CO.trend model. 

# Model diagnostics for NO2.trend.seasonal
```{r}
par(mfrow=c(2,2))
plot(NO2.trend.seasonal, labels.id = NULL)
par(mfrow=c(1,1))
```
Residuals versus fitted plot: Does not violate assumptions. The mean is about zero and there seems to be constant variance.
Q-Q plot: The fit to the line is fairly solid, thus no drastic violation of assumptions. The Q-Q plot could be improved a bit. 
Scale-location: Does not violate assumptions. The mean is about zero and there seems to be constant variance. There are a few outliers.
Residuals versus leverage: No clear influential points with regards to Cook's distance. 

## Part C: Auto-Regressive and Moving Average
#Get the residuals from the CO.trend.seasonal model above and store in e.ts:
```{r}
e.ts.CO<-ts(CO.trend.seasonal$residuals)
```
#Get the residuals from the NO2.trend.seasonal model above and store in e.ts:
```{r}
e.ts.NO2<-ts(NO2.trend.seasonal$residuals)
```
#Plot the residuals for the CO.trend.seasonal model NO2.trend.seasonal
```{r}
plot(e.ts.CO, ylab = "Residuals from CO Model")
plot(e.ts.NO2, ylab = "Residuals from NO2 Model")
```
# Plot the autocorrelation (ACF) and partial autocorrelation (PACF) of the residuals of CO.trend.seasonal
```{r}
par(mfrow=c(1,2))
acf(e.ts.CO, main="ACF of Residuals\nfrom CO.trend.seasonal")
pacf(e.ts.CO,main="PACF of Residuals\nfrom CO.trend.seasonal")
par(mfrow=c(1,1))
```
The ACF plot for the residuals of the CO.trend.seasonal shows sinusoidal decay. However, it cuts off at a very high lag, thus we are going to test various moving average components. 
The PACF plot for the residuals of the CO.trend.seasonal shows sinusoidal decay. However, it cuts off at a very high lag, thus we are going to test various autoregressive components. 
Because the ACF and PACF both show sinusoidal decay, we will also test some ARMA models with both autoregressive and moving average components.
Then we will calculate AIC values to assess several model choices. 

# Do we need to consider a first order difference of our residuals?
```{r}
par(mfrow=c(1,2))
acf(diff(e.ts.CO), main="Diff ACF of Residuals\nfrom CO.trend.seasonal")
pacf(diff(e.ts.CO),main="Diff PACF of Residuals\nfrom CO.trend.seasonal")
par(mfrow=c(1,1))
```
No, we do not need to consider a first order difference the residuals of the CO.trend.seasonal model because the ACF shows sinusoidal decay that does not cut off, and the differentiated ACF does not improve this. Thus, the value of d is 0.

# Plot the autocorrelation (ACF) and partial autocorrelation (PACF) of the residuals of NO2.trend.seasonal
```{r}
par(mfrow=c(1,2))
acf(e.ts.NO2, main="ACF of Residuals\nfrom NO2.trend.seasonal")
pacf(e.ts.NO2,main="PACF of Residuals\nfrom NO2.trend.seasonal")
par(mfrow=c(1,1))
```
The ACF plot for the residuals of the NO2.trend.seasonal shows sinusoidal decay. However, it cuts off at a very high lag, thus we are going to test various moving average components. 
The PACF plot for the residuals of the NO2.trend.seasonal shows sinusoidal decay. However, it cuts off at a very high lag, thus we are going to test various autoregressive components. 
Because the ACF and PACF both show sinusoidal decay, we will also test some ARMA models with both autoregressive and moving average components.
Then we will calculate AIC values to assess several model choices. 

# Do we need to consider a first order difference of our residuals?
```{r}
par(mfrow=c(1,2))
acf(diff(e.ts.NO2), main="Diff ACF of Residuals\nfrom NO2.trend.seasonal")
pacf(diff(e.ts.NO2),main="Diff PACF of Residuals\nfrom NO2.trend.seasonal")
par(mfrow=c(1,1))
```
Both plots show sinusoidal decay, which points to using an ARMA model. 

No, we do not need to consider a first order difference the residuals of the NO2.trend.seasonal model because the ACF shows sinusoidal decay that does not cut off, and the differentiated ACF does not improve this. Thus, the value of d is 0.

# Modeling e.ts.CO 

Now we will try out some models for e.ts.CO using various p and q values

ar(1) p=1
```{r}
CO.ar1 <- arima(e.ts.CO, order=c(1,0,0), include.mean=FALSE)
summary(CO.ar1)
```
AIC = 1423.98

ar(2) p=2
```{r}
CO.ar2 <- arima(e.ts.CO, order=c(2,0,0), include.mean=FALSE)
summary(CO.ar2)
```
AIC = 1421.05

ar(3) p=3
```{r}
CO.ar3 <- arima(e.ts.CO, order=c(3,0,0), include.mean=FALSE)
summary(CO.ar3)
```
AIC = 1416.17

ma(1) p=0, q=1
```{r}
CO.ma1 <- arima(e.ts.CO, order=c(0,0,1), include.mean=FALSE)
summary(CO.ma1)
```
AIC = 1449.39

ma(2) p=0, q=2
```{r}
CO.ma2 <- arima(e.ts.CO, order=c(0,0,2), include.mean=FALSE)
summary(CO.ma2)
```
AIC = 1438.97

ma(3) p=0, q=3
```{r}
CO.ma3 <- arima(e.ts.CO, order=c(0,0,3), include.mean=FALSE)
summary(CO.ma3)
```
AIC = 1426.24

arma(1,3) p=1, q=3
```{r}
CO.arma13 <- arima(e.ts.CO, order=c(1,0,3), include.mean=FALSE)
summary(CO.arma13)
```
AIC = 1418.89

arma(1,2) p=1, q=2
```{r}
CO.arma12 <- arima(e.ts.CO, order=c(1,0,2), include.mean=FALSE)
summary(CO.arma12)
```
AIC = 1417.89

arma(2,3) p=2, q=3
```{r}
CO.arma23 <- arima(e.ts.CO, order=c(2,0,3), include.mean=FALSE)
summary(CO.arma23)
```
AIC = 1419.15

Based on the above AIC values, we would choose model AR(3) because it has the lowest value. 
As a final step, we will use the auto.arima function on e.ts.CO. 
```{r}
CO.auto <- auto.arima(e.ts.CO,approximation=FALSE)
summary(CO.auto)
```
The auto arima function supports the use of AR(3) as our model, with an AIC of 1416.27. 
We will move forward using AR(3) to model the residuals of our CO.trend.seasonal model.

# Modeling e.ts.NO2 

Now we will try out some models for e.ts.CO using various p and q values. 

ar(1) p=1
```{r}
e.ts.NO2

NO2.ar1 <- arima(e.ts.NO2, order=c(1,0,0), include.mean=FALSE)
summary(NO2.ar1)
```
AIC = 3766.75

ar(2) p=2
```{r}
NO2.ar2 <- arima(e.ts.NO2, order=c(2,0,0), include.mean=FALSE)
summary(NO2.ar2)
```
AIC = 3757.73

ar(3) p=3
```{r}
NO2.ar3 <- arima(e.ts.NO2, order=c(3,0,0), include.mean=FALSE)
summary(NO2.ar3)
```
AIC = 3746.67

ma(1) p=0, q=1
```{r}
NO2.ma1 <- arima(e.ts.NO2, order=c(0,0,1), include.mean=FALSE)
summary(NO2.ma1)
```
AIC = 3826.81

ma(2) p=0, q=2
```{r}
NO2.ma2 <- arima(e.ts.NO2, order=c(0,0,2), include.mean=FALSE)
summary(NO2.ma2)
```
AIC = 3805.23

ma(3) p=0, q=3
```{r}
NO2.ma3 <- arima(e.ts.NO2, order=c(0,0,3), include.mean=FALSE)
summary(NO2.ma3)
```
AIC = 3775.63.79

arma(1,2) p=1, q=2
```{r}
NO2.arma12 <- arima(e.ts.NO2, order=c(1,0,2), include.mean=FALSE)
summary(NO2.arma12)
```

AIC = 3747.93

arma(1,3) p=1, q=3
```{r}
NO2.arma13 <- arima(e.ts.NO2, order=c(1,0,3), include.mean=FALSE)
summary(NO2.arma13)
```
AIC = 3749.47

arma(2,3) p=2, q=3
```{r}
NO2.arma23 <- arima(e.ts.NO2, order=c(2,0,3), include.mean=FALSE)
summary(NO2.arma23)
```
AIC = 3751.75

Based on the above AIC values, we would choose model AR(3) because it has the lowest AIC value. 
As a final step, we will use the auto.arima function on e.ts.NO2. 
```{r}
NO2.auto <- auto.arima(e.ts.NO2,approximation=FALSE)
summary(NO2.auto)
```
AIC = 3746.13
Because it has a lower AIC (3744.05) than the model from the auto.arima function (3746.02), we will move forward using ARMA(1,3) with a first order difference of the residuals for our NO2.trend.seasonal model.

# Part D: Assessment of Models -- update these
We used AIC and diagnostics to assess the models for CO. 
```{r}
AIC(CO.ar1) 
AIC(CO.ar2)
AIC(CO.ar3)
AIC(CO.ma1)
AIC(CO.ma2)
AIC(CO.ma3) 
AIC(CO.arma12)
AIC(CO.arma13)
AIC(CO.arma23)
AIC(CO.auto)
```
The lowest AIC is the CO.ar3, which is what the auto.arima function produced as well. 
Therefore the model we would choose is AR(3). 

We also used AIC and diagnostics to assess the models for N02. 
```{r}
AIC(NO2.ar1)
AIC(NO2.ar2) 
AIC(NO2.ar3)
AIC(NO2.ma1)
AIC(NO2.ma2)
AIC(NO2.ma3) 
AIC(NO2.arma12) 
AIC(NO2.arma13)
AIC(NO2.arma23)
AIC(NO2.auto)
```
The lowest AIC is the NO2.arma(13), which is what the auto.arima function produced as well. 
Therefore the model we would choose is ARMA(1,3). 

Now we will consider diagnostics of the best CO model and second best CO model according to their AIC.
```{r}
tsdiag(CO.arma13, lag = 30)
```
The standardized residuals graph shows that there is noise in the ARMA(1,3) model of the residuals.
The ACF of residuals shows no significant lags in the ARMA(1,3) model.
Finally, the Ljung-Box statistic shows that at lag 10, the pvalue is very, very close to 0. While it is still adequate for lags 1-9, we could do better.

```{r}
tsdiag(CO.ar3, lag = 30)
```
The standardized residuals graph shows that there is noise in the AR(3) model of the residuals.
The ACF of residuals shows no significant lags in the AR(3) model.
Finally, the Ljung-Box statistic shows that the p-values do not touch 0, and is at adequate at all the lags.
All of these attributes from the diagnotics plots indicate a valid model.

Now we will consider diagnostics of the best NO2 model and second best NO2 model according to their AIC.
```{r}
tsdiag(NO2.arma23, lag = 30)
```
The standardized residuals graph shows that there is noise in the ARMA(2,3) model of the residuals.
The ACF of residuals shows no significant lags in the ARMA(2,3) model.
Finally, the Ljung-Box statistic shows that the p-values do not touch 0, and is at adequate at all the lags.
All of these attributes from the diagnotics plots indicate a valid model. Now, let's look at the model with the best AIC, AR(3).
```{r}
tsdiag(NO2.ar3, lag = 30)
```
The standardized residuals graph shows that there is noise in the AR(3) model of the residuals.
The ACF of residuals shows no significant lags in the AR(3) model.
Finally, the Ljung-Box statistic shows that the p-values do not touch 0, and is at adequate at all the lags.
All of these attributes from the diagnotics plots indicate a valid model. Because this model also had the lowest AIC, we will use it moving forward.

# Part E: Diagnostics
Regarding the diagnostics of the models we have chosen, there does not appear to be any concern that would warrant a transformation, such as a log transform. 

### Part 2: Building Multivariate Time Series Models

## Part A: Seasonality
Same as part 1a

## Part B: Trends
Same as part 1b

## Part C: Auto-Regressive and Moving Average
```{r}
allResiduals <- data.frame(e.ts.CO, e.ts.NO2)
colnames(allResiduals) <- c("CO","NO2")
cor(allResiduals)
```
Correlation between residuals of NO2 and CO is 0.58914.

# Build VARMA model to CO and NO2 residuals
```{r, include=FALSE}
AICmatrix <- matrix(NA, 3, 4)
for(p in 1:3){
  for(q in 0:3){
    varma.model <- VARMACpp(allResiduals, p=p, q=q, include.mean=F)
    AICmatrix[p,q+1] <- varma.model$aic
  }
}
```
## Part D: Assessment of Models

We will analyze the AICmatrix to find which model has the lowest AIC. 
```{r}
AICmatrix
```
According to the AICmatrix, the model with p=2 and q=2 has the lowest AIC, this we should use these values to build our model. 
The AIC for p=2 and q=2 is 7.16333. The next best model is p=3 and q=3, with an AIC of 7.1883 . 
We will build these 2 models and compare them using diagnostics.  
```{r setup, include = FALSE}
varma.model2 <- VARMACpp(allResiduals, p=3, q=3, include.mean=F)
varma.model2
MTSdiag(varma.model2)
```
The CCF plots shows the correlations are all within the dashed line, meaning that the correlations are not statistically different from 0.
The significance plot of CCM shows a few lags that are below the dashed line. While this is not too problematic, we can do better.
The plot of p-values for Ljung-Box statistics shows that the model is adequate until around lag 12.
The residual plot for CO and NO2 show noise, *yay!* 

```{r}
varma.model <- VARMACpp(allResiduals, p=2, q=2, include.mean=F)
varma.model
MTSdiag(varma.model)
```
The CCF plots shows the correlations are all within the dashed line, meaning that the correlations are not statistically different from 0.
The significance plot of CCM shows a few lags that are below the dashed line. While this is not too problematic, we can do better.
The plot of p-values for Ljung-Box statistics shows that the model is adequate until around lag 11.
The residual plot for CO and NO2 show noise, *yay again!*

After investigating the AIC values and the diagnostic plots of the two comparable VARMA models, we have decided to use VARMA(2,2) due to its lowest AIC value and acceptable diagnostic plots. 

## Part E: Diagnostics
While we considered the diagnostics to be acceptable, they could be better. For the p-values for Ljung-Box statistic plot, it would be better if there were more adequate lags (i.e. p-values do not become significant until future lags). In the significance plot of CCM, there were a few points below the dashed line. However, these observations do not pose any major concerns for the remainder of our analysis.

### Part 3: Simulating from Univariate and Multivariate Time Series Models
```{r}
next.year.time <- c(1:(365))
next.year <- data.frame(time = next.year.time)

COmean <- predict(CO.trend.seasonal, newdata = next.year)
NO2mean <- predict(NO2.trend.seasonal, newdata = next.year)

set.seed(10)
T.simUCO = arima.sim(CO.ar3$model,365)
set.seed(5)
T.simUNO2 = arima.sim(n=365, list(ar=c(NO2.ar3$coef[1],NO2.ar3$coef[2],NO2.ar3$coef[3])),sd=sqrt(NO2.ar3$sigma2))

```

```{r}
T.simM = VARMAsim(365,phi=varma.model$Phi,theta=varma.model$Theta,sigma=varma.model$Sigma)
```
## Part A: Ability to reproduce appearance
Multivariate and Univariate: CO 
```{r}
{plot(COmean + T.simUCO)
lines(dailyAQ$CO.GT.[1:365] + allResiduals$CO[1:365],col="blue")
lines(COmean + T.simM$series[,1], col = "red")}
```
Above is a plot of ? which shows ___

Multivariate and Univariate: NO2
``` {r}
{plot(NO2mean + T.simUNO2)
lines(dailyAQ$NO2.GT.[1:365] + allResiduals$NO2[1:365],col="blue")
lines(NO2mean + T.simM$series[,2], col = "red")}
#plot(dailyAQ$NO2.GT.[1:365] + allResiduals$NO2[1:365],col="green")
#plot(NO2mean + T.simUNO2)
#plot(NO2mean)
plot(T.simUNO2)
```


## Part B: Ability to reproduce observed trends

Univariate: CO
```{r}
COsim<-COmean+T.simUCO
CO.trend.seasonal.sim <- lm(COsim[next.year.time] ~ next.year.time + sin(2*pi*next.year.time/7) + cos(2*pi*next.year.time/7))
summary(CO.trend.seasonal)
summary(CO.trend.seasonal.sim)
```
The coefficient estimates for time and the seasonality components are very similar to each other for each respective model. This means means that using our simulated data, we were able to reproduce the observed trends of each time series.

Univariate: NO2
```{r}
NO2sim<-NO2mean+T.simUNO2
NO2.trend.seasonal.sim <- lm(NO2sim[next.year.time] ~ next.year.time + sin(2*pi*next.year.time/7) + cos(2*pi*next.year.time/7))
summary(NO2.trend.seasonal)
summary(NO2.trend.seasonal.sim)
```
The coefficient estimates for time and the seasonality components are very similar to each other for each respective model. This means means that using our simulated data, we were able to reproduce the observed trends of each time series.

Multivariate: CO
```{r}
MCOsim<-COmean+T.simM$series[,1]
MCO.trend.seasonal.sim <- lm(MCOsim[next.year.time] ~ next.year.time + sin(2*pi*next.year.time/7) + cos(2*pi*next.year.time/7))
summary(CO.trend.seasonal)
summary(MCO.trend.seasonal.sim)
```
The coefficient estimates for time and the seasonality components are fairly similar to each other for each respective model. The seasonality coefficients are slightly different, but still close to each other. This means that using our simulated data, we were able to reproduce the observed trends of each time series.

Multivariate: NO2
```{r}
MNO2sim<-NO2mean+T.simM$series[,2]
MNO2.trend.seasonal.sim <- lm(MNO2sim[next.year.time] ~ next.year.time + sin(2*pi*next.year.time/7) + cos(2*pi*next.year.time/7))
summary(NO2.trend.seasonal)
summary(MNO2.trend.seasonal.sim)
```
The coefficient estimates for time and the seasonality components are fairly similar to each other for each respective model. The intercept (107.88 vs 114.92) and seasonality coefficients are slightly different, but still close to each other. This means that using our simulated data, we were able to reproduce the observed trends of each time series fairly well.

## Part C: Ability to reproduce seasonality

Univariate: CO
```{r}
pg.COsim <- spec.pgram(COsim,spans=9,demean=T,log='no')
```
Univariate: NO2
```{r}
pg.NO2sim <- spec.pgram(NO2sim,spans=9,demean=T,log='no')
```
this is weird, ask about in office hours

Multivariate: CO
```{r}
pg.MCOsim <- spec.pgram(MCOsim,spans=9,demean=T,log='no')
```
Multivariate: CO
```{r}
pg.MNO2sim <- spec.pgram(MNO2sim,spans=9,demean=T,log='no')
```

All of the above simulated periodograms for univariate and multivariate models all show a peak frequency at around 0.14, similar to the observed periodograms. This means we were able to accurately reproduce seasonality of each time series.

## Part D: Ability to reproduce observed mean and variance

Univariate: CO
```{r}
mean(dailyAQ$CO.GT.)
mean(COsim)
var(dailyAQ$CO.GT.)
var(COsim)
```
The means are very close to each other (4.365 vs. 4.278), which indicates our model reproduces the observed mean well. However, the variances are slightly different (3.623 vs 1.4479). This coincides with the fact that our simulated values were all slightly lower than each of the observed value.

Univariate: NO2
```{r}
mean(dailyAQ$NO2.GT.)
mean(NO2sim)
var(dailyAQ$NO2.GT.)
var(NO2sim)
```
The means are very close to each other (164.5335 vs. 162.2001), which indicates our model reproduces the observed mean well. However, the variances are rather different (2657.72 vs 880.8054). ??????????

Multivariate: CO
```{r}
mean(dailyAQ$CO.GT.)
mean(MCOsim)
var(dailyAQ$CO.GT.)
var(MCOsim)
```
The means are very close to each other (4.365 vs. 4.3988), which indicates our multivariate model reproduces the observed mean for CO well. The variances are also similar (3.623 vs 2.847), indicating that our multivariate model also reproduces the observed variance CO well.

Multivariate: NO2
```{r}
mean(dailyAQ$NO2.GT.)
mean(MNO2sim)
var(dailyAQ$NO2.GT.)
var(MNO2sim)
```
The means are very close to each other (164.5335 vs. 160.9884), which indicates our multivariate model reproduces the observed mean for NO2 well. The variances are also fairly similar (2657.722 vs 2056.203), indicating that our multivariate model also reproduces the observed variance for NO2 well.

## Part E: Ability to reproduce auto-correlation

Univariate: CO
```{r}
par(mfrow=c(1,2))
acf(CO.ts)
acf(COsim)
par(mfrow=c(1,1))
par(mfrow=c(1,2))
pacf(CO.ts)
pacf(COsim)
par(mfrow=c(1,1))
```

Univariate: NO2
```{r}
par(mfrow=c(1,2))
acf(NO2.ts)
acf(NO2sim)
par(mfrow=c(1,1))
par(mfrow=c(1,2))
pacf(NO2.ts)
pacf(NO2sim)
par(mfrow=c(1,1))
```

Multivariate: CO
```{r}
par(mfrow=c(1,2))
acf(CO.ts)
acf(MCOsim)
par(mfrow=c(1,1))

par(mfrow=c(1,2))
pacf(CO.ts)
pacf(MCOsim)
par(mfrow=c(1,1))
```
The PACFs are pretty different...ask in office hours


Multivariate: NO2
```{r}
par(mfrow=c(1,2))
acf(NO2.ts)
acf(MNO2sim)
par(mfrow=c(1,1))
par(mfrow=c(1,2))
pacf(NO2.ts)
pacf(MNO2sim)
par(mfrow=c(1,1))
```

## Part F: Ability to reproduce observed cross-correlation
```{r}
cor(dailyAQ$CO.GT.,dailyAQ$NO2.GT.) #observed
cor(COsim,NO2sim) #simulated from univariate
cor(MCOsim,MNO2sim) #simulated from multivariate
```

### Bonus

```{r}
CO.forecast <- forecast(e.Tmin.lm, h=6)
```