---
title: "Project 1"
author: "Shayne Cassidy (sc5za) Sarah Winston Nathan (swn2bf) Sarah Nelson (skn5mq)"
date: "10/20/2019"
output: html_notebook
---

```{r setup, include=FALSE}
require("knitr")
#traindir <- "C:/Users/swnathan/Documents/4y1s/sys4021/InClassR/Data/TrainData"
#sourcedir <-"C:/Users/swnathan/Documents/4y1s/sys4021/InClassR/RCode"
#traindir <- "/Users/shaynecassidy/Desktop/4021/Data/TrainData"
#sourcedir <-"/Users/shaynecassidy/Desktop/4021/RCode"
traindir <- "/Users/student/SYS4021/Data/Trains"
sourcedir <- "/Users/student/SYS4021/RCode"
opts_knit$set(root.dir = sourcedir)
opts_chunk$set(warning=FALSE)
```

Load data and combine data from 2001-2018 into totacts.
```{r}
# Source AccidentInput
setwd("/Users/shaynecassidy/Desktop/4021/RCode")
source("AccidentInput.R")

# list of data frames for each year of accident data
acts <- file.inputl(traindir)

# data frame with all accidents from all years from 2001 - 2018

# Get a common set the variables
comvar <- intersect(colnames(acts[[1]]), colnames(acts[[8]]))

# the combined data frame
totacts <- combine.data(acts, comvar)
```


Create a new variable, Casualty, where Casualty = TOTINJ + TOTKLD
```{r}
totacts$Casualty <- totacts$TOTKLD + totacts$TOTINJ
```

Build a data frame with only extreme accidents for ACCDMG
```{r}
dmgbox <- boxplot(totacts$ACCDMG)
xdmg <- totacts[totacts$ACCDMG > dmgbox$stats[5],]
# Remove duplicates from xdmg and call new data frame xdmgnd
xdmgnd <- xdmg[!(duplicated(xdmg[, c("INCDTNO", "YEAR", "MONTH", "DAY", "TIMEHR",
                                     "TIMEMIN")])),]
```

Build a data frame with only extreme accidents for Casualty
```{r}
casbox <- boxplot(totacts$Casualty)
cas <- totacts[totacts$Casualty > casbox$stats[5],]
# Remove duplicates from cdmg and call new data frame casnd
casnd <- cas[!(duplicated(cas[, c("INCDTNO", "YEAR", "MONTH", "DAY", "TIMEHR",
                                     "TIMEMIN")])),]
```

Treatment of categorical variables

These variables will need to be tested in our hypotheses. So, in order to appropriately test the hypotheses relating to these categorical variables, these categorical variables are recoded into two levels each. 
```{r}
# Create a freight train variable
Freight <- rep(0, nrow(xdmgnd))
Freight[which(xdmgnd$TYPEQ == 1)] <- 1
Freight <- as.factor(Freight)

# Create a human factors variable
HF <- rep(0, nrow(xdmgnd))
HF[which(substr(xdmgnd$CAUSE, 1, 1) == "H")] <- 1
HF <- as.factor(HF)
```

```{r}
# Create a derailment variable
Derail <- rep(0, nrow(casnd))
Derail[which(casnd$TYPE == 1)] <- 1 
Derail <- as.factor(Derail)

# Create a highway-rail crossing variable
HRX <- rep(0,nrow(casnd))
HRX[which(casnd$TYPE == 7)] <- 1
HRX <- as.factor(HRX)
```


## Part 1: Generating Hypotheses
ACCDMG: We used the following visualizations, summary statistics, and external supporting evidence to arrive at our hypotheses for ACCDMG. 
```{r}
# Freight, Human Factors
mean(xdmgnd$ACCDMG)
max(xdmgnd$ACCDMG)

xdmgnd$TYPEQ <- as.numeric(xdmgnd$TYPEQ)
xdmgnd$TYPEQ <- factor(xdmgnd$TYPEQ, labels = c("Freight", "Passenger", "Commuter", "Work",  "Single", "CutofCars", "Yard", "Light", "Maint"))
barplot(table(xdmgnd$TYPEQ))
bwplot(as.factor(TYPEQ)~ACCDMG, data = xdmgnd, main = "Box Plots of Accident Damage", xlab = "Damage ($)", ylab = "Type Train")

xdmgnd$Cause <- rep(NA, nrow(xdmgnd))
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "M")] <- "M" 
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "T")] <- "T"
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "S")] <- "S"
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "H")] <- "H"
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "E")] <- "E"
xdmgnd$Cause <- factor(xdmgnd$Cause) 
barplot(table(xdmgnd$Cause))
bwplot(as.factor(Cause)~ACCDMG, data = xdmgnd, main = "Box Plots of Accident Damage", xlab = "Damage ($)", ylab = "Cause")

```

Casualties: We used the following visualizations, summary statistics, and external supporting evidence to arrive at our hypotheses for Casualties.
```{r}
# Derailment, Hwy Crossing

casnd$TYPE <- factor(casnd$TYPE, labels = c("Derailment", "HeadOn", "Rearend", "Side", "Raking", "BrokenTrain", "Hwy-Rail", "GradeX", "Obstruction", "Explosive", "Fire","Other","SeeNarrative" ))
barplot(table(casnd$TYPE))

mean(casnd$Casualty)
max(casnd$Casualty)
plot(1:18, tapply(casnd$Casualty, casnd$YEAR, max), type = "l", xlab = "Year", ylab = "Number Casualties", main = "Max Casualties per Year")

```

Based on the above analysis, we have been able to form some well-informed hypotheses from the visualizations. They are as follows, listed as null and alternative hypotheses and followed by how we arrived at them:

..As for accident severity in regards to **Accident Damage (ACCDMG)**:
*1) H0: Accidents involving freight trains do not significantly increase accident damage.*
*Ha: Accidents involving freight trains do significantly increase accident damage.*

*This is an actionable hypothesis because more regulations can be put in place to promote safe driving with freight trains in dangerous areas.*

*2) H0: Accidents caused by Human Factors do not significantly increase accident damage.*
*Ha Accidents caused by Human Factors do significantly increase accident damage.*

*This is an actionable hypothesis because better recurrent training programs can be implemented to improve conductor and engineer abilities to safely manage tasks when in control of a train.*

..As for accident severity in regards to **Casualties**: 

*1) H0: Derailment caused accidents do not significantly increase the number of casualties.*  
*Ha: Derailment caused accidents do significantly increase the number of casualties.* 

*This is an actionable hypothesis because train wheels and tracks can be manufactured in order to prevent derailments.*

*2) H0: Highway-rail crossing accidents do not significantly increase number of casualties.*
*Ha: Highway-rail crossing accidents are do significantly increase the number of casualties.*

*This is actionable because if the null hypothesis is accepted, then trains can be required to follow more safety precautions when traveling near highways, and other traffic laws can be put in place to encourage safe driving around highway-railroad crossings. *

## Part 2: ACCDMG Analysis

*Hypothesis 1*

Fill in here steps a-e
(a) The feature and model selection techniques you used to find appropriate models for this problem;
(b) Your treatment of ordinal and categorical variables (i.e., how were
they coded);
(c) How you assessed your models (e.g., adjusted R2, AIC, etc.);
(d) How you diagnosed problems with the models; and
(e) How you adjusted the models based on these assessments.

```{r}
source("SPM_Panel.R")
uva.pairs(xdmgnd[,c("ACCDMG", "TRNSPD", "CARS", "TRKDNSTY", "TONS", "HEADEND1", "CDTRHR", "ENGRS")])
```

Based on the visualizations and correlation values from the scatter plot matices above, we have decided to include TRNSPD, TRKDNSTY, CARS, TONS, and HEADEND1 as quantitative variables in our model.

To test our hypotheses, we will include the two treated categorical variables, Freight and HF (Human Factors), with the selected quantitative variables in our linear model.
```{r}
xdmgnd.lm1 <- lm(ACCDMG ~ Freight + HF  + CARS + TRNSPD + HEADEND1 + TONS, data=xdmgnd)
summary(xdmgnd.lm1)
```

*Hypothesis 2*

Fill in here steps a-e
(a) The feature and model selection techniques you used to find appropriate models for this problem;
(b) Your treatment of ordinal and categorical variables (i.e., how were
they coded);
(c) How you assessed your models (e.g., adjusted R2, AIC, etc.);
(d) How you diagnosed problems with the models; and
(e) How you adjusted the models based on these assessments.


## Part 3: Casualties

*Hypothesis 1*

*1) H0: Derailment caused accidents do not increase the number of casualties*  
*Ha: Derailment caused accidents do increase the number of casualties* 

Fill in here steps a-e
(a) The feature and model selection techniques you used to find appropriate models for this problem;

```{r}
source("SPM_Panel.R")
uva.pairs(casnd[,c("Casualty", "CARS", "TIMEHR", "TEMP", "TRKDNSTY", "TRNSPD", "TONS")])
```
Due to the low correlations between casualty and TIMEHR, TONS, and TRKDNSTY, we will not include these variables in our model. 

```{r}
casnd.lm1 <- lm(Casualty ~ Derail + HRX + TEMP + CARS + TRNSPD, data=casnd)
summary(casnd.lm1)
```


#Plot diagnostic graphs individually
```{r}
plot(casnd.lm1,which=1) #Residual vs. Fitted
```
This plot violates the assumptions for the residual vs fitted diagnostic plot because there is not a mean of zero and there is not a constant variance. 

```{r}
plot(casnd.lm1,which=2) #QQ
```
This graph shows a violation of assumptions because the points clearly do not follow the line. This represents a failure of the Gaussian assumption for the error term. 

```{r}
plot(casnd.lm1,which=3) #Scale-Location
```
This graph shows a violation of assumptions because the points do not have a mean of zero and there is not a constant variance. 

```{r}
plot(casnd.lm1,labels.id = NULL, which=4) #Cook's distance
```

```{r}
plot(casnd.lm1,which=5) #Redisuals vs. Leverage
```
These graphs shows a violation of assumptions because there are incidents with very large Cook's distances.

The analysis above shows that this model does not meet the assumptions required for assessment of this linear model. So, we can conclude that this is not a good model for assessing casualties from train accidents. Therefore, we next perform a log transormation of this model to assess if this creates a more effective model of casualties. 

```{r}
casnd.lm2 <- lm(log(Casualty) ~ Derail + HRX + TEMP + CARS + TRNSPD, data=casnd)
summary(casnd.lm2)
```

#Plot diagnostic graphs individually
```{r}
plot(casnd.lm2,which=1) #Residual vs. Fitted
```
This plot violates the assumptions for the residual vs fitted diagnostic plot because there is not a mean of zero and there is not a constant variance. 

```{r}
plot(casnd.lm2,which=2) #QQ
```
This graph shows a violation of assumptions because the points clearly do not follow the line. This represents a failure of the Gaussian assumption for the error term. 

```{r}
plot(casnd.lm2,which=3) #Scale-Location
```
This graph shows a violation of assumptions because the points do not have a mean of zero and there is not a constant variance. 

```{r}
plot(casnd.lm2,labels.id = NULL, which=4) #Cook's distance
```

```{r}
plot(casnd.lm2,which=5) #Redisuals vs. Leverage
```
Although these graphs still show a violation of assumptions due to large Cook's distances, there is a slight improvement from the previous model after taking the log transformation. 

```{r}
summary(casnd.lm1)
summary(casnd.lm2)
#AIC
AIC(casnd.lm1)
AIC(casnd.lm2)
#BIC
AIC(casnd.lm1,k=log(nrow(casnd)))
AIC(casnd.lm2,k=log(nrow(casnd)))


```
The adjusted R2 value for the first model is 0.02212 and for the transformed model it is 0.06468. While the transformed model shows a larger adjusted R2 value and lower AIC and BIC values than the first model, none of these values indicate a strong model. This indicates that neither model is appropriate to assess both hypotheses related to casualties. Thus, in order to determine an effective model to predict casualty severity, we would like to investigate non-linear regression models. 

Although the p-values for Derail (0.002514, 0.026724) and HRX (3.73e-05, < 2e-16) technically indicate significance, the assumptions for the linear models were not met, thus we cannot draw conclusions about our hypotheses from these models. 



(b) Your treatment of ordinal and categorical variables (i.e., how were
they coded);
(c) How you assessed your models (e.g., adjusted R2, AIC, etc.);
(d) How you diagnosed problems with the models; and
(e) How you adjusted the models based on these assessments.


*Hypothesis 2* 

Fill in here steps a-e
(a) The feature and model selection techniques you used to find appropriate models for this problem;
(b) Your treatment of ordinal and categorical variables (i.e., how were
they coded);
(c) How you assessed your models (e.g., adjusted R2, AIC, etc.);
(d) How you diagnosed problems with the models; and
(e) How you adjusted the models based on these assessments.


## Part 4: Evidence and Recommendations to FRA

### Part a)

### Part b)